\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{tikz}

\geometry{a4paper, margin=1in}

\title{Visualization of Sound Spectra of a Whistle}
\author{Jorge R. Gherson}

\begin{document}

\maketitle
\newpage 

\tableofcontents
\newpage

\section{General Objective}
We have a dataset containing the sound spectra of a whistle collected at different mass 
flow rates, pressures, and frequencies. The main objective of this project is to 
visualize the sound spectra in a way that would allow a Convolution Neural Network (CNN) 
or a human to efficiently see the relationship between the sound spectra and the 
mass flow rate. 

\input{exploratory-analysis.tex}

\newpage

\section{Mathematical Formulation}
In order to make the relationship easily visible for a Convolution 
Neural Network, I will carry out a mathematical definition for how 
the network transforms the raw acoustic spectrum into a mass flow rate prediction. 

Let our normalized acoustic spectrum for a single flow rate be 
defined as a one-dimensional vector 

\begin{equation}
        \mathbf{x} = \begin{bmatrix}
                x_1 & x_2 & \cdots & x_N
        \end{bmatrix}^T
\end{equation}

where \(N\) is the total number of frequency bins. The core operation of 
a CNN is the discrete 1D convolution, which applies a set of \(K\) 
trainable kernels (i.e., filters) to the input to detect features like 
whistle peaks or harmonic intervals. 

Let a single filter be a vector \(\mathbf{w}^{(k)}\) of length \(M\). 
The convolution operation computes a new sequence \(\mathbf{z}^{(k)}\) 
for this filter. The \(i\)-th position is given by: 

\begin{equation}
        z_i^{(k)} = \sum_{j=1}^{M} \mathbf{x}_{i+j-1}\cdot \mathbf{w}_j^{(k)} + b^{(k)}
\end{equation}

where \(\mathbf{w}_j^{(k)}\) represents the weights of the \(k\)-th filter, and
\(b^{(k)}\) the bias of the \(k\)-th filter. This represents a mathematical 
cross-correlation. When the shape of the filter matches the shape of a peak 
in the spectrum, \(z_i\) spikes. To allow the network to learn complex, non-linear physical relationships, 
we apply a Rectified Linear Unit (ReLU) to the output of the convolution: 

\begin{equation}
        h_i^{(k)} = \max{(0, z_i^{(k)})}.
\end{equation}

Then, pooling reduces the size of the data, making the model computationally efficient and ``translation invariant.''
The maximum pooling takes the maximum value over a sliding window of size \(P\): 

\begin{equation}
        p_i^{(k)} = \max{h_{P(j-1)+1}^{(k)}, \cdots, h_{Pj}^{(k)}}.
\end{equation}

Later in the network, we use global average pooling (GAP), which takes the average of the entire remaining sequence 
of length \(L\) for each filter \(k\), resulting in a single scalar value per filter: 

\begin{equation}
        g^{(k)} = \frac{1}{L}\sum_{i=1}^L h_i^{(k)},
\end{equation}

which flattens the matrix into a vector \(\mathbf{g} = \begin{bmatrix} g^{(1)} & g^{(2)} & \cdots & g^{(K)}\end{bmatrix}^T\). 
Once the spatial features (i.e., peaks, harmonics) are extracted and flattened into the vector \(\mathbf{g}\), a standard 
neural network layer combines them to predict the flow rate. Let \(\mathbf{W}_d\) be the weight matrix of the dense layer and 
\(b_d\) be the bias vector of the dense layer. Then, 

\begin{equation}
        \mathbf{d} = \max{(0, W_{d\mathbf{g}}+b_d)}.
\end{equation}

Since we are predicting a continuous physical variable, the final layer is a single node with a linear activation function: 

\begin{equation}
        \hat{y} = \mathbf{w}_{\text{out}}^T d + b_{\text{out}}
\end{equation}

where \(\hat{y}\) represents the predicted mass flow rate \(y\) and \(\mathbf{w}_{\text{out}}\) represents the final weight vector mapping 
the learned features to the prediction. However, in order to train a network efficiently, we must also define how ``wrong'' our predictions can be 
compared to the actual mass flow rate. For regressions, we usually use Mean Squared Error (MSE) over a batch of \(B\) training samples: 

\begin{equation}
        L = \frac{1}{B}\sum_{i=1}^B (\hat{y}_i - y_i)^2
\end{equation}

% \newpage 
% \input{mechanics.tex}

\newpage

\input{software-implementation.tex}


\end{document}